---
output:
  github_document:
    toc: true
    html_preview: false
---
  
```{r setup, include = FALSE}
  knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/REAsDME-",
  out.width = "100%"
)
```

<!-- README.md is generated from README.Rmd. Please edit that file -->

# foot: An R package for processing building footprints
*[WorldPop Research Group, University of Southampton](https://www.worldpop.org/)*

The `foot` package is designed to provide a set of consistent and flexible tools
for processing 2D vector representations of buildings. The functionality
includes basic geometry and morphology measures, distance and clustering
metrics. These calculations are supported with helper functions for spatial
intersections and tiled reading/writing of data. Two main wrapper functions
provide standardised workflows for process building polygons to summarise
metrics within polygons or on a gridded surface.

## Installation
The `foot` package can be installed directly from Github.
```r
devtools::install_github("wgpg/foot", auth_token="tkn", build_vignettes=TRUE)
```
where, "tkn" is an authentication token. This is required because `foot` is
still in a private repo. You can obtain a token from
[https://github.com/settings/tokens](https://github.com/settings/tokens) and
generating a personal token with 'repo' level access. This will be a long
character string.

Alternatively, clone the repository and install using the code in using the code
in `./pkg_build.R`

Note that building and running the code may require additional packages:
`stars`, `raster`, `sf`, `data.table`, `lwgeom`, `purrr`. et al.

## Quick Start
A sample dataset of building footprints is provided: 
```
# load the sample
data("kampala")

# 2D vector building polygons
kampala$buildings
```

Read the vignette on basic usage `vignette(footsteps)`. For more advanced uses
and creating gridded data layers, see `vignette(bigfoot)`.

### Basic Usage
```{r, message=FALSE, warning=FALSE}
library(foot)
# load sample data
data("kampala")
buildings <- kampala$buildings
zones <- kampala$adminZones
grid <- kampala$mastergrid
```

The `foot` package provides tools to calculate and summarise building morphology
measures at multiple scales. These include
```{r message=FALSE, warning=FALSE, tidy=FALSE}
# building-level metrics
buildings$built_area <- fs_area(buildings, 
                                unit="m^2")
  head(buildings)
```

```{r, message=FALSE, warning=FALSE, tidy=FALSE}
# Area-level summary metrics
# index the buildings to zones
building_zone <- zonalIndex(buildings, 
                            zones, 
                            zoneField = "Id", 
                            returnObject = TRUE)

# summarise within small areal units
admin_area <- fs_area_mean(building_zone, 
                           index="Id", 
                           unit="m^2")
  head(admin_area)
```

```{r message=FALSE, warning=FALSE, tidy=FALSE}
# calculated along a raster within a focal window
gridded <- calculate_bigfoot(buildings, 
                             metrics="area_mean", 
                             focalRadius=200,
                             template=grid)

  raster::plot(raster::raster(gridded))
  plot(sf::st_geometry(buildings), add=TRUE)
```


### Outputs
Rasters (or tables):  
  
1. Binary settlement indicators
2. Counts of structures
3. Building area (total, mean, median, standard dev., coeff. var.)
4. Building perimeter (total, mean, median, standard dev., coeff. var.)
5. Nearest neighbour distance (mean, median, standard dev.)
6. Nearest neighbour index
7. Structure orientation angle (normalised entropy)
8. Compactness (Polsby-Popper) (mean, median)
9. Roundess (mean, median)

A full list of function names can be retrieved with `get_fs_metrics()`.

## Contributions
Contributions are welcome. Raise or respond to an issue, or create a new branch
to develop a feature/modification and submit a pull request.

## Repository Structure
The repository is structured as an R package with an additional folder "wd" that
is a working directory for storing scripts, input data, and output data.

**./pkg_build.R**  
A script to build the R package and install it on your machine. Optionally,
builds vignettes.

**./data/**
Folder containing internal data files and sample building footprints for the
vignettes.

**./data-raw/**
Folder containing the script to create the internal data files.

**./doc/** 
A folder containing the markdown and scripts for the vignettes.

**./man/**  
A folder containing function documentation created by Roxygen. **Do not edit
these files.** Instead, use Roxygen to document each function (see
`./R/footFun.R` for example) and build the documentation using
`devtools::document()` (see `./pkg_build.R` for example).

**./Meta/**
A folder created as part of building vignettes.

**./R/**  
A folder containing functions for the R package. Each file contains a function
with Roxygen documentation for the function at the top of the script.

**./wd/code/**  
A folder containing scripts. See the example script `./wd/code/example_script.R`
for a template. Note that the vignette scripts and .html source have been copied
here for easy access for users who don't want to install/build the package.

**./wd/in/**  
A folder containing input data. This folder is included in the .gitignore for
the repository, so files you save here will not be uploaded to github or shared
with collaborators on the repository.

**./wd/out/**  
A folder containing outputs from the scripts. This folder is included in the
.gitignore for the repository, so files you save here will not be uploaded to
github or shared with collaborators on the repository.

