
```{r setup, include = FALSE}
  knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/REAsDME-",
  out.width = "100%"
)
```

<!-- README.md is generated from README.Rmd. Please edit that file -->

# foot: An R package for processing building footprints
*[WorldPop Research Group, University of Southampton](https://www.worldpop.org/)*

While remote sensing has long been used to monitor urbanisation patterns, in
recent years there has been an increasing availability in finer resolution
satellite data covering large areas of the globe. This very high resolution
imagery (often <1 m spatial resolution), combined with increased computing power
is producing new datasets on urban areas. In particular, machine learning
algorithms are being applied to detect, automatically extract, and map full sets
of building features in a scene. These automated methods add to the manually
digitised information such as from [OpenStreetMap](http://www.openstreetmap.org)
and the property datasets available from some city governments.

Such building footprint datasets provide a new level of detail on urban areas,
particularly in places which might otherwise lack detailed maps on cities and
rapidly growing areas. Despite their spatial detail, building footprints
typically lack any other attribute information to help differentiate land uses
or other neighbourhood characteristics. However, the size, shape, orientation,
and clustering of structures produces a spatial pattern that can suggest areas
of different land use or activities.

The `foot` package is designed to provide a set of consistent and flexible tools
for processing 2D vector representations of buildings. The functionality
includes basic geometry and morphology measures, distance and clustering
metrics. These calculations are supported with helper functions for spatial
intersections and tiled reading/writing of data. 

## Installation

The `foot` package can be installed directly from Github.
```r
devtools::install_github("wgpg/foot", auth_token="tkn", build_vignettes=TRUE)
```
where, "tkn" is an authentication token. This is required because `foot` is
still in a private repo. You can obtain a token from
[https://github.com/settings/tokens](https://github.com/settings/tokens) and
generating a personal token with 'repo' level access. This will be a long
character string.

Note that building and running the code may require additional packages:
`stars`, `raster`, `sf`, `data.table`, `lwgeom`, `mmap`. et al.

## Quick Start

A sample dataset of building footprints is provided: 
```
# load the sample
data("kampala", package="foot")

# 2D vector building polygons
kampala$buildings
```

### Vignettes

Vignettes are provided as an introduction to `foot`. The vignette on basic usage
is available using `vignette("footsteps", package="foot")`. The supplied
datasets can be used to replicate this vignette. For more advanced uses and for
a discussion on creating gridded data layers, see `vignette("bigfoot",package="foot")`.


### Basic Usage

```{r, message=FALSE, warning=FALSE}
library(foot)

# load sample data
data("kampala", package="foot")
buildings <- kampala$buildings
zones <- kampala$adminZones
grid <- kampala$mastergrid
```

The `foot` package provides tools to calculate and summarise building morphology
measures at multiple scales. These include building-level measures.
```{r message=FALSE, warning=FALSE, tidy=FALSE}
# building-level metrics
buildings$built_area <- fs_area(buildings, 
                                unit="m^2")
  head(buildings)
```

As well as area-level summaries.
```{r, message=FALSE, warning=FALSE, tidy=FALSE}
# Area-level summary metrics
# index the buildings to zones
building_zone <- zonalIndex(buildings, 
                            zones, 
                            zoneField = "Id", 
                            returnObject = TRUE)

# summarise within small areal units
admin_area <- fs_area_mean(building_zone, 
                           index="Id", 
                           unit="m^2")
  head(admin_area)
```

Or gridded summary outputs, with the options to include a circular focal window.
```{r message=FALSE, warning=FALSE, tidy=FALSE}
# calculated along a raster within a focal window
gridded <- calculate_bigfoot(buildings, 
                             metrics="area_mean", 
                             focalRadius=200,
                             template=grid)

  raster::plot(raster::raster(gridded))
  plot(sf::st_geometry(buildings), add=TRUE)
```


### Outputs

Rasters in GeoTiff format (or tables):  
  
1. Binary settlement indicators
2. Counts of structures
3. Building area (total, mean, median, min, max, standard dev., coeff. var.)
4. Building perimeter (total, mean, median, min, max, standard dev., coeff. var.)
5. Nearest neighbour distance (mean, median, standard dev.)
6. Nearest neighbour index
7. Structure orientation angle (normalised entropy)
8. Compactness (Polsby-Popper) (mean, median)
9. Roundess (mean, median)

A full list of function names can be retrieved with `foot::get_fs_metrics()`.


## Contributions

Contributions are welcome. Raise or respond to an issue, or create a new branch
to develop a feature/modification and submit a pull request.


## Acknowledgements

This work was undertaken by members of the WorldPop Research Group at the
University of Southampton. Funding support comes from the Bill and Melinda Gates
Foundation and the United Kingdom Foreign, Commonwealth & Development Office as
part of the Geo-Referenced Infrastructure and Demographic Data for Development
project (GRID3) (OPP1182408). Project partners in [GRID3](https://grid3.org/)
include the WorldPop Research Group, the United Nations Population Fund, the
Flowminder Foundation, and the Center for International Earth Science
Information Network within the Earth Institute at Columbia University.
