---
title: "Gridded footprint calculation layers"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
description: >
  Building footprints or polygons representing the outline shape of structures
  can provide a unique data source for studying urban areas at high spatial
  resolutions. This vignette will demonstrate how to use the `foot` package to
  produce high-spatial resolution gridded data layers of geometric and
  morphological measures, potentially for large regions.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gridded footprint calculation layers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy.opts=list(width.cutoff=80), 
  tidy=TRUE
)
```

The `foot` package was developed by WorldPop at the University of Southampton
([www.worldpop.org](https://www.worldpop.org)) to support geometric calculations
and zonal summaries of measures from building footprint polygons. The
`vignette("footsteps", package="foot")` provides an introduction to package and 
the functionality of calculating and summarising morphology measures. This 
vignette builds on those methods and demonstrates a more advanced workflow to 
produce gridded summaries of buildings measures. In particular, the focus is on 
using `foot::calculate_bigfoot` in order how to handle very large (i.e. 
national-scale) datasets of building footprints.

This vignette use a portion of Ordnance Survey's OS OpenMap Local data of
building footprint polygons. The polygons cover the area of Southampton. These
data were originally processed and made available by Alasdair Rae
([http://ajrae.staff.shef.ac.uk/buildings/](http://ajrae.staff.shef.ac.uk/buildings/))
under the [OS OpenData License](http://os.uk/opendata/licence). The data contain
Ordnance Survey data © Crown copyright and database right 2018.

```{r setup}
library(foot)
```

## Calculations with `foot`

The central function for producing gridded layers is `calculate_bigfoot`. It is
designed to support processing large sets of building polygons into gridded
(GeoTiff) summary layers. The procedure is intended to be more memory-efficient
by splitting the processing area into "tiles" which can then be handled
individually or in parallel. This function works as a wrapper to
`calculate_footstats` as well as several helper functions within the `foot`
package for managing the input/output and creating and indexing spatial zones.

By default the function performs calculations in parallel (which can be changed
with the argument `parallel=FALSE`. Users can adjust the number of computing 
cores used with the argument `nCores=`) To monitor the supplied values and
processing steps, set `verbose=TRUE`.

### Main inputs

Users need to supply `calculate_bigfoot()` with:

* A path to a file of building footprints in a spatial vector data format (e.g.
`.gpkg`, `.shp`)
* A filepath to a template gridded dataset specifying the extent and resolution
for the outputs (e.g. `.tif`)

For example:

```{r}
# local file paths used for demonstration
dir <- "/home/jochem/Documents/projects/0_data"

bldgPath <- file.path(dir, "soton_buildings.gpkg")
sotonGrid <- file.path(dir, "soton_borough_100m.tif")
```

While `R` objects can be supplied for these parameters, it is recommended to
supply character strings of paths to the files. `calculate_bigfoot` will only
read the subset of data needed for a particular processing step. This is
intended to allow users to process much larger datasets than can be held in
memory.

### Basic calculations

With the key inputs for the file paths set, the processing can be executed with
a single function.

```{r}
# basic function call with default parameters
calculate_bigfoot(X=bldgPath,
                  template=sotonGrid,
                  metrics="settled",
                  verbose=TRUE)
```

### Specifying outputs

By default the outputs of `bigfoot` are saved as GeoTiffs in `R`'s `tempdir()`
which is changed for the each session. Each grid is named by the short name of
the calculated metric.

```{r Fig1, fig.cap="Binary settled raster for Southampton at 100m resolution", fig.height=6, fig.width=6}
# retrieve the gridded outputs
outGrid <- raster::raster(file.path(tempdir(), "settled.tif"))

raster::plot(outGrid)
```

Users can specify an output path to another folder location. Additionally a
"tag" can be specified as a parameter to the function. The tag is appended to
the beginning of each file name. This can be useful for identifying different
outputs. These options are demonstrated in the code block below.

### Multiple metrics

As with `calculate_footstats` multiple metrics and summary statistics can be
supplied to `bigfoot` as a vector of strings. Partiular, shortened versions of 
the metric names should be used. The list of metrics is available
in `foot::list_fs()` or it can be retrieved with several look-up
functions (see `?get_fs_metrics`).

```{r}
# get list of available metric names in shortened format
list_fs("short_name")

# re-running the basic call
calculate_bigfoot(X=bldgPath,
                  metrics=c("settled","area_cv"),  # multiple metrics
                  template=sotonGrid, 
                  outputPath=tempdir(),  # output folder location as a string
                  outputTag="soton")  # filename tag
```
```{r Fig2, fig.cap="Coeff. of variation in building area for Southampton at 100m resolution", fig.height=6, fig.width=6}
# retrieve the gridded outputs
# must add the "tag" to the filename
outGrid <- raster::raster(file.path(tempdir(), "soton_area_cv.tif"))

raster::plot(outGrid)
```

### Focal window statistics

In the examples above, the footprint statistics are summarised for buildings
which intersect the pixels of the template grid. Internally this is handled by
`zonalIndex()`. However, it is also possible to calculate and summarise building
features within a local, circular window. The output is still associated with
each template grid cell, similar to a moving window analysis. The focal radius
distance is always specified in meters.

```{r}
# moving focal window calculations
calculate_bigfoot(X=bldgPath,
                  metrics=c("settled",
                            "area_mean",
                            "perim_total"),
                  focalRadius=300,  # buffer radius in meters
                  template=sotonGrid, 
                  outputPath=tempdir(), 
                  outputTag="soton")
```

Note that when a focal radius is specified, this value is always appended to the
end of the file names so that the outputs can be identified.
```{r Fig3, fig.cap="Total building perimeter in a 300m radius window. Output at a 100m resolution", fig.height=6, fig.width=6}
# note that the filename includes both a tag and a focal radius value
outGrid <- raster::raster(file.path(tempdir(), "soton_perim_total_300.tif"))

raster::plot(outGrid)
```


## Options and finer control

The `calculate_bigfoot` function is set up with default values that should work
under most conditions; however, there is additional flexibility for users to
specify alternative parameters. 

### Specifying geometry units

To override the default units used in the geometry calculations, a named list of
unit strings can be supplied to the `controlUnits` argument. This list can
contain named items for `areaUnit`, `perimUnit`, and `distUnit`. The value of
each item should be coercible with `units::as_units`.

```{r Fig4, fig.cap="Total building perimeter in KM", fig.height=6, fig.width=6}
# change the default units used to calculate area and distance
calculate_bigfoot(X=bldgPath,
                  metrics=c("area_mean",
                            "perim_total"),  
                  controlUnits=list(areaUnit="m^2",  # change default units
                                    perimUnit="km"),
                  template=sotonGrid, 
                  outputPath=tempdir(),  
                  outputTag="soton",
                  parallel=FALSE,
                  verbose=TRUE)  

# plot the total perimeter, measured in kilometres
outGrid <- raster::raster(file.path(tempdir(), "soton_perim_total.tif"))
raster::plot(outGrid)
```
In the above example, note that the units for the nearest neighbour distance
were not specified in the `controlUnits` list. The default value was still used.

### Filtering buildings

In some settings it may be preferable to exclude very small and/or very large
building footprint polygons. The lower and upper bounds for filtering can be
specified with `minArea` and `maxArea` arguments. The values for these filters
are in the same units specified by `controlUnits` or the default value for area
calculations. Note that an "area" footprint statistic does not need to be
requested.

```{r}
calculate_bigfoot(X=bldgPath,
                  metrics=c("shape_mean",
                            "count",
                            "perim_total"),  
                  controlUnits=list(areaUnit="m^2"),
                  minArea=50,  # footprints must be larger than 50 m^2
                  maxArea=1000,  # footprints must be smaller than 1000 m^2
                  template=sotonGrid, 
                  outputPath=tempdir(),  
                  outputTag="soton",
                  parallel=FALSE,
                  verbose=TRUE)  
```

In the map of the results, note the much smaller number of structures in pixels
around the central business districts and southwestern edge of the study region.
```{r Fig5, fig.cap="Count of buildings with area >50 m^2 and <1000 m^2", fig.height=6, fig.width=6}
outGrid <- raster::raster(file.path(tempdir(), "soton_count.tif"))

raster::plot(outGrid)
```

### Tile size

The size of the processing tiles, specified in pixel dimensions (rows, columns)
can be an important factor in the efficiency of the calculations. Smaller tile
regions result in fewer building footprints being read/processed at one time,
but there is an overhead computational cost of reading/writing files. The
default value is 500 pixels. For the small demonstration shown here that results
in one tile for the whole region. To show multiple tile processing, a small size
is supplied and the processing is done in parallel with verbose output.

```{r}
calculate_bigfoot(X=bldgPath,
                  template=sotonGrid,
                  metrics=c("settled","compact_mean"),
                  tileSize=c(100, 75),  # rows x columns in pixels
                  parallel=FALSE,
                  verbose=TRUE)
```

***
```{r}
sessionInfo()
```

