---
title: "Using custom summary functions in `foot`"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
description: >
  Building footprints or polygons representing the outline shape of structures
  can provide a unique data source for studying urban areas at high spatial
  resolutions. This vignette will demonstrate how to supply custom functions to
  the `foot` package to extend the range of potential summary metrics.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gridded footprint calculation layers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy.opts=list(width.cutoff=80), 
  tidy=TRUE
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `foot` package was developed by WorldPop at the University of Southampton
([www.worldpop.org](https://www.worldpop.org)) to support geometric calculations
and summaries of measures from building footprint polygons. This vignette
demonstrates how users can extend the basic functionality of
`calculate_footstats` and `calculate_bigfoot` by making and supplying their own
functions to summarise footprint characteristics. For an introduction to the
package, see `vignette("footsteps")`.

```{r setup}
library(foot)
```

## Calculations with `foot`

As noted in the introductory vignettes, `foot` primarily uses `calculate_foots`
to calculate and summarise metrics. Internally this function uses `data.table`
in order to handle large sets of building footprints and efficiently summarise
them. The attributes to be summarised (`what`) are supplied to function names
(`how`). These internal structures also allow for user-defined functions to be
specified.

### Data preparation

To demonstrate using custom functions, we will first add some additional "data"
to the footprints which we will use.

```{r}
data("kampala", package="foot")

buildings <- kampala$buildings
adminzones <- kampala$adminZones

# Adding random data
# categorical variable
buildings$category <- sample(LETTERS[1:5], size=nrow(buildings), replace=T)
# continuous variable
buildings$mult <- sample(rnorm(nrow(buildings), mean=10, sd=2))
```

We can use any attributes of the footprints within `calculate_footstats` and
`calculate_bigfoot`, not only the built-in morphology measures listed by
`list_fs(what='all')`.

```{r, tidy=F}
# summarising a new data value
calculate_footstats(buildings,
                    adminzones,
                    what="mult", # new attribute to summarise
                    how="mean",
                    verbose=F)
```

### Additional built-in functions

The internal `foot` functions are documented in `?fs_functions`; however, these
functions are intended to be used within the wrapper functions of `foot` and are
rarely intended to be used as standalone functions. One built-in summary
function, not applied by default to any characteristics, is `majority`. It is
designed to summarise categorical data. This function is available for users in
the same manner of specifying the `how` argument.

```{r, tidy=FALSE}
# get the majority category in each zone
calculate_footstats(buildings, 
                    adminzones, 
                    what="category", 
                    how="majority", 
                    verbose=F)
```

The `majority` function is similar to the idea of supplying a user-defined
function which is demonstrated in the next section.

## User-defined summary functions

Creating functions for use with `foot` follows the same procedures and syntax
for functions in `R` in general. They must be declared with `<- function()` and
they must be available within the environment where `foot` functions are being
used. They should return a single, summary value since the function will be
applied to footprints by zone.

The name of the function is what is passed to `foot` as an argument to `how`.
The argument(s) to the custom function can be named anything, but they will
typically be values within the footprints to be summarised.

The example below shows a simple function that calculates the sum of the square
root of the values. We will apply it to 'area', and `foot` will automatically
pre-calculate this characteristic since it is not present in the column names of
the footprints.

```{r, tidy=FALSE}
# simple function example 1
f1 <- function(v){
  units(v) <- NULL # ignore units in our function
  return(sum(sqrt(v)))
}

# applying custom summary function to area
calculate_footstats(buildings,
                    adminzones,
                    what="area", how="f1",
                    verbose=F)
```

Although this function was just used to process area, the function can be used
for any continuous value. It can also be used on multiple characteristics or
combined with other lists of functions, just like any other built-in function in
`foot`.

```{r, tidy=FALSE}
calculate_footstats(buildings,
                    adminzones,
                    what=list("area","perimeter"), how="f1",
                    verbose=F)
```

## Functions with multiple arguments

In some instances, a custom function may need to make use of two or more
characteristics from within the building footprint datasets. The built-in
functions in `foot` are primarily designed to work with a single value (e.g.
area or perimeter).

While it may often be quicker to pre-calculate the combination, it could be
advantageous to use a function. To make sure multiple attributes are supplied to
the summary function, the arguments in `what` should be specified using a
special type (`fs_varlist`). The `fs_varlist` creates a nested list within the
processing to keep the argument together. Keep in mind that the arguments are
passed by **position** not be name, so the order must match what the function is
expecting.

### Creating a Perimeter/Area ratio

An example of a custom function using two characteristics is the average
perimeter-area ratio. We can compare this to the built-in function which uses a
Polsby-Popper metric (`fs_compact`).

```{r, tidy=FALSE}
# average perimeter-area ratio
pa <- function(p, a){
  return(mean(p / a))
}

# used to summarise within zones
# note that fs_varlist is still within a list
calculate_footstats(buildings,
                    adminzones,
                    what=list(list("compact"), fs_varlist("area","perimeter")),
                    how=list(list("mean"), list("pa")),
                    verbose=T
                   )
```

## Accessing `R` objects other than the footprints

A more complicated scenario exists when a user-defined function needs to access
data which is not an attribute of the footprints dataset. In order to access the
non-footprint data, a partial function must be created and supplied.

In the example below, a simple constant value is supplied to a summary function;
however, the idea extends to any object. This process is how the nearest
neighbour index is calculated in `foot` by drawing on the spatial zones object
as well as the footprints.
```{r, tidy=F}
# external "data"
d1 <- 0.001

# This will NOT work because argument 'd' is missing
# f2 <- function(x, d){
#   return(sum(d * x))
# }
# 
# calculate_footstats(buildings, adminzones, what="area", how="f2", verbose=T)

# Instead...
# example of creating a partial function
gen_f3 <- function(d){
  force(d) # must include
  function(x){
    return(sum(d * x))
  }
}

# generate the function and initialise it with `d1` from above.
f3 <- gen_f3(d1)

# this now uses the generated function, and `d` is found
calculate_footstats(buildings, 
                    adminzones, 
                    what="area", 
                    how="f3", 
                    verbose=F)
```

In this vignette, the `foot` package has been extended to incorporate
user-defined functions. These functions can use one or more values from within
the footprints, or even access other objects in the environment. While the
examples used `calculate_footstats`, the same approaches can be used to create
new gridded summary metrics with `calculate_bigfoot`.

***
```{r}
sessionInfo()
```
