---
title: "Basic building footprint calculations"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
description: >
  Building footprints or polygons representing the outline shape of structures
  can provide a unique data source for studying urban areas at high spatial
  resolutions. This vignette will introduce the `foot` package and demonstrate
  some of the tools provided to calculate summary measures of geometric
  properties and to produce tabular and gridded outputs.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic building footprint calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy.opts=list(width.cutoff=80), 
  tidy=TRUE
)
```

While remote sensing has long been used to monitor urbanisation patterns, in
recent years there has been an increasing availability in finer resolution
satellite data covering large areas of the globe. This very high resolution
imagery (often <1 m spatial resolution), combined with increased computing power
is producing new datasets on urban areas. In particular, machine learning
algorithms are being applied to detect, automatically extract, and map full
sets of building features in a scene. Such building footprint datasets provide a
new level of detail on urban areas, particularly in places which might otherwise
lack detailed maps on cities and rapidly growing areas. Despite their spatial
detail, building footprints typically lack any other attribute information to
help differentiate land uses or other neighbourhood characteristics. However,
the size, shape, orientation, and clustering of structures produces a spatial
pattern that can suggest areas of different land use or activities.

```{r setup}
library(foot)
```

The `foot` package was developed by WorldPop at the University of Southampton
([www.worldpop.org](https://www.worldpop.org)) to support geometric calculations
and zonal summaries of measures from building footprint polygons. This vignette
will demonstrate some of the core functionality in the package, including:

* The available measurements and summary statistics
* How to define different types of spatial zones for summaries
* Example workflows to produce tabular and gridded output

To demonstrate the package, this vignette will use a supplied sample of building
footprint polygons produced by Microsoft Bing Maps
([Source](https://github.com/microsoft/Uganda-Tanzania-Building-Footprints))
from an area in Kampala, Uganda. These footprints have been reprocessed into a
spatial data format which can be read with `sf`.

## Load the sample dataset
```{r}
data("kampala")

buildings <- kampala$buildings
adminzones <- kampala$adminZones
clusters <- kampala$clusters
```

The sample dataset is a list of four named objects: 

* "buildings" - polygons of building footprints in `sf` format. Contains 8480
records. 
* "mastergrid" - geoTiff `RasterLayer` aligned to WorldPop datalayers. This will
be used as a template for producing other gridded outputs
* "adminZones" - 34 polygons in `sf` format for zonal statistics 
* "clusters" - 10 small polygons in `sf` format for sample sites

Note that the adminZones and cluster boundaries are purely artificial and were
created for demonstration purposes only.

For more information, see `?kampala`.

```{r echo=FALSE, fig.cap="Sample buildings and zones in Kampala", fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(sf::st_geometry(buildings), 
     col=sf::sf.colors(12, categorical=TRUE),
     border=NA, axes=TRUE)

plot(sf::st_geometry(adminzones), border="red", add=T)
```

## Calculations with `foot`
All function names for calculating metrics begin with `fs_` (short for
"footprint statistic"). The second part of the name includes the group of
measures (e.g. area or perimeter), and finally an abbreviation for the summary
statistic (e.g. 'cv' for coefficient of variation or 'sd' for standard
deviation).

### Available metrics and summary measures
Currently implemented are measures for:

* presence (binary and count)
* area
* perimeter
* Euclidean nearest neighbour distance
* nearest neighbour index
* angle of rotation
* compactness and shape indices

A table of metrics is available in `foot::fs_footprint_metrics`.

## Basic usage
The simplest usage of `foot` involves supplying a set of building footprints to
the desired metric function.

```{r}
# mean of footprint areas
built_area <- fs_area_mean(buildings)

built_area

# standard deviation of perimeters
perim_sd <- fs_perim_sd(buildings)

perim_sd
```

With no other argument supplied, all the footprints are treated as coming from
the same zone for the summary. Default unit values are used for geometry
measurements. All functions return the same format - a `data.table` with a
column for the index and the named column for the summarised footprint metric.

There is also an option to supply pre-calculated column to each function with
the `col=` argument. If a pre-calculated geometry value exists, that field can
be used or, if it is absent, the appropriate geometries needed for the metric
will be calculated.

All functions follow the same parameter template, and, because the first
argument is always data, all the functions allow for "piping" syntax.

```{r, tidy=F}
# random sample of buildings
result <- buildings %>%
          dplyr::filter(FID_1 %in% sample(FID_1, 1000)) %>%
          fs_area_mean()

# show the resulting table
result
```

Most low-level geometry calculations rely on `sf` and `lwgeom` and are not
exposed in `foot`. A few exceptions are for calculating the nearest neighbour
distance, the minimum bounding circle, and the rotated minimum bounding
rectangle.

### Nearest neighbour distances
Distances can be calculated between footprints (as points or polygons), between
footprints and other objects, and within a maximum search distance.

```{r}
# nearest distance for the first 10 buildings
# measured between polygon edges
fs_NNdist(buildings[1:10,], buildings, maxSearch=200, unit="m")

# omitting 'Y' measures distance among the footprints
fs_NNdist(buildings[1:10,], maxSearch=NULL)  # unrestricted distance
```

Note that distance calculations are slower for polygons and for unprojected
coordinate systems. It is recommended that a maximum search radius is always
used. Internally the calculations are done with a `data.table` to benefit from
multithreading capabilities.

### Rotation angles
A somewhat less conventional geometric measure is derived from the rotated
bounding rectange. This is the rectangle enclosing a polygon (or set of points)
which has been rotated to minimise the area. In contrast, a bounding box is
always oriented along the x and y axes.

```{r}
# To obtain the rotated rectangle as a spatial object
mbr <- fs_mbr(buildings[5,], returnShape=T)

plot(sf::st_geometry(buildings[5,]))
plot(mbr, border="red", add=T)

# Or obtain just the angle
fs_mbr(buildings[5,], returnShape=F)

```

The angles can be summarised as an entropy measure and normalised to describe
how much the angles of structures depart from a regular grid pattern (available
in `fs_angle_entropy()`).


## Creating and supplying zone indices
Rather than treating all features as a single summary zone, it's more common to
want to summarise metrics within smaller areas. There are several ways to supply
information on the zones.

### Index by vector
A vector of indices for a zone can be supplied to the functions as a column
name, column number, or passed as a standalone vector object. The length of a
vector of indices must match the number of building polygons.

```{r}
# create a vector of ten random zones
id <- sample(1:10, nrow(buildings), replace=T)
buildings$id <- id   # add it to the buildings object
  table(buildings$id)  # splitting observations into 10 groups

# pass the index by name
fs_area_total(buildings, index="id")  # note default units

# pass the index by number
fs_count(buildings, index=3)

# pass a separate vector object
print(fs_angle_entropy(buildings, index=id, normalize=T))
```

### Index by shape
Rather than supplying a pre-calculated column or vector of zonal indices,
buildings can be assigned a zone based on a spatial join. This returns a field
called "zoneID" which other calculation functions can use.

```{r message=FALSE, warning=FALSE}
# Return a table of index values
zID <- zonalIndex(buildings, adminzones, returnObject=F)
  zID # the xID column are row numbers
  
# Alternatively, join the zones to create a new spatial object
# A column with a custom zone can be added
zObj <- zonalIndex(buildings, clusters, zoneField="Id", returnObject=T)
  zObj
```

```{r fig.height=6, fig.width=6}
# use the new object
zarea <- fs_area_mean(zObj, index="Id")  
clusters <- merge(clusters, zarea, by.x="Id", by.y="index")
  plot(clusters["fs_area_ha_mean"])
```

The `zonalIndex` function works by spatial intersection. This produces some
(potentially uesful) side effects to be aware of.

The default behavior is for any intersecting structure to included in the zone.
```{r warning=FALSE}
# Note the selected structures extend beyond the cluster boundary
plot(sf::st_geometry(clusters)[[6]])
plot(sf::st_geometry(buildings), add=T)
plot(sf::st_geometry(zObj[zObj$zoneID==6,]), col="red", add=T)
```

Alternatively, the intersection can return a clipped set of buildings.
```{r message=FALSE, warning=FALSE}
zClip <- zonalIndex(buildings, clusters, zoneField="Id", clip=T)

plot(sf::st_geometry(clusters)[[6]])
plot(sf::st_geometry(buildings), add=T)
plot(sf::st_geometry(zClip[zClip$Id==6,]), col="red", add=T)
```

This clips the footprints via intersection, potentially leaving small slivers of
structures in the zone which will affect the feature statistics.

An additional side effect of the intersection is that overlapping zones are
allowed and this duplicates and associates the footprints in both zones.

```{r message=FALSE, warning=FALSE}
# create a temporary shape by shifting one cluster
newClusters <- st_sfc(sf::st_geometry(clusters)[[1]], 
                      sf::st_cast(sf::st_geometry(clusters)[[1]] + c(.001,.0001), "POLYGON"),
                      crs=sf::st_crs(clusters)
                     )

newClusters <- st_sf(geometry=newClusters, crs=sf::st_crs(clusters))

newObj <- zonalIndex(buildings, newClusters, clip=T)

# areas of overlap are in a purple hue
plot(sf::st_geometry(newClusters))
plot(sf::st_geometry(newObj[newObj$zoneID==1,]), col="red", add=T)
plot(sf::st_geometry(newObj[newObj$zoneID==2,]), col=sf::sf.colors(n=1, alpha = 0.5), add=T)
plot(sf::st_geometry(buildings), add=T)
```

These side effects are allowed because they allow for flexibility to support
types of "focal" summaries of statistics and to produce a true gridded measure
of footprint metrics.


## Calculating multiple metrics
The `calculate_footstats()` function provides a convenient wrapper to the
individual footprint statistic functions and other helper functions. The
function accepts a variety of input formats (see `?calculate_footstats`).
Multiple metrics can be calculated for the same sets of buildings and zones.

```{r, tidy=F}
# Creates zonal index and calculates multiple metrics
results <- calculate_footstats(buildings, 
                               adminzones, 
                               metrics=c("fs_area_mean","fs_area_cv"), 
                               gridded=F)
  print(results)  
```

Note that this function uses the default arguments and units for creating zones
and calculating metrics. In the above results the standard deviation was added
because it is required for the CV calculation. Also note that one zone has only 
one structure in it, so NA values are returned for the standard deviation and 
CV.

The argument `metrics` will accept a string or a vector of strings for specific
metric names. Users may also supply `"ALL"` or `"NODIST"` to calculate all
available metrics or all bar the nearest neighbour distance-related ones,
respectively.


## Gridded outputs 
Often the desired output is a gridded dataset of a measure, rather than
summarising metrics to polygon zones. A gridded dataset can be achieved with the
`foot` functions in several ways.

(@) The simplest method is to attach buildings (and summary measures) to grid
cells by the centroid point location. A fast approach is to use cell numbers.
```{r, fig.width=6, fig.height=6, tidy=F}
mgrid <- kampala$mastergrid # ~100m raster layer

# get the cell numbers from the grid
cellIDs <- raster::cellFromXY(mgrid, 
                              sf::st_coordinates(sf::st_centroid(buildings)))
# use these numbers as the zone
gridRes <- calculate_footstats(buildings, 
                               index=cellIDs, 
                               metrics=c("fs_perim_total",
                                         "fs_area_total",
                                         "fs_count",
                                         "fs_settled"),
                               gridded=F)
# buildings lacking a grid cell are NA
gridRes <- na.omit(gridRes)
  gridRes
  
# create a blank raster and fill with summary measures
outSettled <- mgrid
outSettled[] <- NA
# `$index` are the cell numbers from the raster 
outSettled[gridRes$index] <- gridRes$fs_area_ha_total

outCount <- mgrid
outCount[] <- NA
outCount[gridRes$index] <- gridRes$fs_count

raster::plot(outCount)
plot(sf::st_geometry(buildings), add=T)
```

(@) Pixel-level calculation. While centroid-based calculations are fast,
sometimes it may be necessary to know the true amount of settlement *per pixel*.
The zonal index functions can be used to intersect buildings to grid squares,
and then supply those "pixels" as zones to the feature calculations as shown.
```{r message=FALSE, warning=FALSE, tidy=F}
# convert the grid to a `stars` object and then `sf` to get polygons
sgrid <- stars::st_as_stars(mgrid)
sgrid <- sf::st_as_sf(sgrid)

# add a cell number ID
sgrid$cellIDs <- raster::cellFromXY(mgrid, sf::st_coordinates(sf::st_centroid(sgrid)))

# create a zonal index on the building objects
pxZone <- zonalIndex(buildings, 
                     sgrid, 
                     "cellIDs", 
                     returnObject=T, 
                     clip=T)

pxResult <- calculate_footstats(pxZone, 
                                index="cellIDs",
                                metrics=c("fs_area_total", "fs_settled"))
  print(pxResult)
```

As before, the metrics are linked to the grid by their cell numbers. The results
can also be merged to the `stars` or `sf` representation of the grid and quickly
rasterized based on geometry. The difference between using centroids or pixels
is most visible with large structures.
```{r, fig.cap="Centroid-based (left) and a pixel-based (right) map of total building area", fig.show='hold', fig.width=6}
# create output
pxSettled <- mgrid
pxSettled[] <- NA
pxSettled[pxResult$index] <- pxResult$fs_area_ha_total

# compare centroid and pixel settled maps
par(mfrow=c(1,2))
raster::plot(outSettled, zlim=c(0,1), xlim=c(32.623, 32.626), ylim=c(0.338, 0.342))
plot(sf::st_geometry(buildings), add=T)
raster::plot(pxSettled, zlim=c(0,1), xlim=c(32.623, 32.626), ylim=c(0.338, 0.342))
plot(sf::st_geometry(buildings), add=T)
```

(@) Write gridded data to GeoTiff files. Using `calculate_footstats` a template
raster and an output location can be specified. The template can be a `raster`
or `stars` object. Each metric will be written to a separate grid. This approach
also demonstrates how an index parameter of `sf` type can be provided directly
to the wrapper function.
```{r, fig.width=6, fig.height=6, tidy=F}
outputPath <- tempdir()

# re-using the sf pixels as "zones"
calculate_footstats(buildings, 
                    index=sgrid, 
                    metrics=c("fs_area_mean", "fs_settled"),
                    gridded=TRUE, 
                    template=mgrid,  # using a rasterlayer
                    outputPath=outputPath # or a temp directory is created
                   )

# get the output
raster::plot(raster::raster(paste0(outputPath, "/fs_area_ha_mean.tif")))
plot(st_geometry(buildings), add=T)
```

The above approaches work fine for small study areas, but may be too slow over
large areas and with many footprint features. Producing country-scale gridded
datasets with tools from `foot` is addressed in the second vignette ("bigfoot").


***
```{r}
sessionInfo()
```

