---
title: "Basic building footprint calculations"
description: >
  Building footprints or polygons representing the outline shape of structures
  can provide a unique data source for studying urban at high spatial
  resolutions. This vignette will introduce the foot package and demonstrate
  some of the tools provided to calculate summary measures of geometric
  properties and to produce tabular and gridded outputs.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic building footprint calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

While remote sensing has long been used to monitor urbanisation patterns, in
recent years there has been an increasing availability in finer resolutions
atellite data covering large areas of the globe. This very high resolution
imagery (often <1 m spatial resolution, combined with increased computing power
is producing new datasets on urban areas. In particular machine learning
algorithms are being applied to detect and automatically extract and map full
sets of building features in a scene. Such building footprint datasets provide a
new level of detail on urban areas, particularly in places which might otherwise
lack detailed maps on cities and rapidly growing areas. Despite their spatial
detail, building footprints typically lack any other attribute information to
help differentiate land uses or other neighbourhood characteristics. However,
the size, shape, orientation, and clustering of structures produces a spatial
pattern that can suggest areas of different land use or activities.

```{r setup}
library(foot)
```

The `foot` package was developed by WorldPop at the University of Southampton to
support geometric calculations and zonal summaries of measures from building
footprint polygons. This vignette will demonstrate some of the core
functionality in the package, including:

* The available measurements and summary statistics
* How to define different types of spatial zones for summaries
* An example workflow to produce tabular and gridded output

To demonstrate the package, this vignette will use a supplied sample of building
footprint polygons produced by Microsoft Bing Maps
([Source](https://github.com/microsoft/Uganda-Tanzania-Building-Footprints))
from an area in Kampala, Uganda.

## Load the sample dataset
```{r}
data("kampala")

buildings <- kampala$buildings
adminzones <- kampala$adminZones
clusters <- kampala$clusters
```






