---
title: "Create gridded building metric covariates"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
description: >
  Building footprints or polygons representing the outline shape of structures
  can provide a unique data source for studying urban at high spatial
  resolutions. This vignette will introduce a workflow using the `foot` package
  and the tools provided to produce gridded covariates of footprint metrics for
  large regions.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create gridded building metric covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## TO-DO
* move to a vignette? How to include/access large set of footprints?
* alternative: create a separate building dataset from MS buildings
* re-write and add parallel looping code for tiles
* check mdsumner's grout package for GDAL-native tiles


## Demonstration workflow
```{r setup}
library(foot)
library(stars)   # for reading/writing some GDAL files
library(raster)  # to demonstrate with raster tools

# Package to download WorldPop mastergrids
# devtools::install_github("wgpg/wpgpDownloadR")
library(wpgpDownloadR) 
```

This demonstration will use a set of building footprints produced by Ecopia in
partnership with Maxar (Digitize Africa data © 2020 Maxar Technologies,
Ecopia.AI). These cover South Sudan and have been reprocessed to merge into a
single Geopackage file in WGS84 projection (4326).
```{r}
# path to building footprint data
bldgPath <- "//worldpop.files.soton.ac.uk/worldpop/Projects/WP517763_GRID3/Working/SSD/sett/made_2/ssd_buildings_wgs84_merge.gpkg"
```


The processing will create an output that aligns with the WorldPop mastergrid.
Each pixel will also become a processing location to summarise the footprint
metrics.
```{r, tidy=F}
# load the mastergrid
mgPath <- wpgpGetCountryDataset(ISO3="SSD", 
                                covariate="level0_100m_2000_2020")
mgrid <- read_stars(mgPath, proxy=TRUE)
  
  print(mgrid)
```
Note that the proxy setting for `stars` does not bring the file into memory, but 
provides access to the files's dimensions.

### Create processing tiles
The large extent of the mastergrid is split into roughly equal sized tiles.
These tiles can then be used to process the footprint metrics in parallel across
multiple cores or multiple machines. If the metrics require a local summary,
such as within a buffer around the processing location, then the tiles need a
buffer (or "halo") as well.

```{r, tidy=F}
# create tiles
tiles <- gridTiles(mgrid, 
                   px=c(200, 200), # approximate size row, col pixels
                   overlap=0
                  )

# create buffered tiles (assuming max of 250m buffer at each location)
tilesBuff <- gridTiles(mgrid,
                       px=c(200,200),
                       overlap=3  # overlap in pixels, at least 250m
                      )

  head(tiles)
```

In an actual process, each row in the table of tiles serves a "job" to be done 
in parallel. For this demonstration, one tile will be use.
```{r}
i <- 200 # example
```

Example job tile for processing
```{r, tidy=F}
# select the tile
job <- tiles[i,]

# get the tile extent
rio <- list(nXOff = job[[1]], 
            nYOff = job[[2]], 
            nXSize = job[[5]], 
            nYSize = job[[6]], 
            bands = c(1))

# create a tile from the mastergrid
# without a buffer will define the output locations
mgTile <- read_stars(mgPath,
                     RasterIO=rio,
                     proxy=TRUE
                    )

# and create a buffered tile for extracting buildings
job <- tilesBuff[i,]

# get the tile extent
rio <- list(nXOff = job[[1]], 
            nYOff = job[[2]], 
            nXSize = job[[5]], 
            nYSize = job[[6]], 
            bands = c(1))

# in the same way
mgBuffTile <- read_stars(mgPath,
                         RasterIO=rio,
                         proxy=TRUE
                        )
```

The buffered tiles have an extra 3 pixel on each side.
```{r}
dim(mgTile)
dim(mgBuffTile)
```

### Extract buildings
The building footprints should be extracted using the extent of the buffered
tile so that a focal window analysis at each processing location can be
computed. If the computer has enough RAM to read in the full datafile, then the 
spatial filter step can be skipped and more quickly subset/cropped with `[`.
```{r}
# set-up a spatial filter for file reading
# based on the bounding box of the tile
wkt <- st_as_text(st_as_sfc(st_bbox(mgBuffTile)))

# read in the footprints
buildings <- st_read(bldgPath,
                     wkt_filter=wkt
                    )
  buildings
```

